A Data-Driven Framework for Neural Field Modeling

### 2011-01-16 ltx/first_order_framework/response_to_reviews.tex

\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{soul}
\usepackage{anysize} %<-for setting margins
\marginsize {1.6cm}{1.6cm}{2.0cm}{2cm}
\usepackage{color}
\newcommand{\dean}[1]{\textcolor{red}{#1}}
\newcommand{\mike}[1]{\textcolor{green}{#1}}
\newcommand{\parham}[1]{\textcolor{blue}{#1}}

\title{Responses - A Data-Driven Framework for Neural Field Modeling}
\author{D. Freestone, P. Aram, M. Dewar, K. Scerri, D. Grayden, V. Kadirkamanathan}

\begin{document}
    \maketitle

    The authors would very much like to thank the reviewers for their encouraging comments and their efforts to help to improve this paper. Please find below the authors responses to the reviewers questions and suggestions regarding the manuscript. 
\\

@@

    \end{enumerate}

    \section{Reviewer 2}

\begin{enumerate}
    \item The handling of the (to borrow a term from numerical weather forecasting) `sub-grid scale physics' is inadequate for nervous systems (see, e.g., equation 8 (now equation 12) [The nonlinear IntegroDifference Equation]). I do very much appreciate the spectral estimation of spatial and temporal scales. But the model errors are lumped into an iid model error term that is not realistic. For instance, the flow of potassium in the extracellular spaces has a known dynamic that will affect network excitability, generating important sub-grid dynamics that is anything but iid. See, for instance, the example in Ullah et al PLoS Comp Biol 2010 where without such dynamics, state tracking and parameter estimation for a neural system fails. A beautiful overview of sub-grid effects in data assimilation is found in Kalnay's 2003 book. I would very much value the authors comments on these issues. I guess were I to respond to this question, that I might consider extending the model error term to an autoregressive model. But I will let the authors answer this question.

	\emph{The authors believe that allowing for time-varying parameters will account for changes in excitability levels. In future work, we will also estimate the disturbance properties (spatial covariance and temporal variance). To address this point in this paper, we have added an extra paragraph to `Neural Field Model' section of the Methods, where we discuss the model mismatch and some basic modeling assumptions, such as stationarity of the parameters. We have also added to the Estimation section, stating that we are assuming the parameters are stationary of the estimation period. In addition, we have added to the discussion (Extensions section) with possible ways to relax these assumptions.}

### 2011-01-17 ltx/first_order_framework/response_to_reviews.tex

\item What is the relation of the substitution of the temporal and spatial basis functions as employed in this paper to Galerkin's projection? How do the two techniques differ? If this is close, the non-local case for a Galerkin projection (typically used in diffusive processes) has not, to my knowledge, been done. Certainly not in a way that is obvious for neural field theory. Perhaps this is my ignorance. This seems important to clarify for the rest of us.

\emph{We have added an extra step in the derivation of the state-space model to highlight the relationship to our basis function decomposition and Galerkin's projection. We have also added an extra Appendix to further explain the relationship.}

\item In the isotropic and translationally invariant case examined here, would a Fourier basis for temporal and spatial structure have been a formally correct alternative?

\emph{Yes. Fourier basis functions would be a correct alternative to the case examined in this paper. In fact, Fourier basis functions or any other orthonormal basis functions would simplify the state-space representation of the model, since $\Gamma$ (defined in equation 21) of the manuscript would be the identity matrix. However, we have chosen Gaussian basis functions due to their semi-compact support. A comment has been added for possible alternative basis functions under equation 22 of the manuscript.}

### 2011-01-17 ltx/first_order_framework/response_to_reviews.tex

    \end{enumerate}

    \section{Reviewer 2}

\begin{enumerate}
    \item The handling of the (to borrow a term from numerical weather forecasting) `sub-grid scale physics' is inadequate for nervous systems (see, e.g., equation 8 (now equation 12) [The nonlinear IntegroDifference Equation]). I do very much appreciate the spectral estimation of spatial and temporal scales. But the model errors are lumped into an iid model error term that is not realistic. For instance, the flow of potassium in the extracellular spaces has a known dynamic that will affect network excitability, generating important sub-grid dynamics that is anything but iid. See, for instance, the example in Ullah et al PLoS Comp Biol 2010 where without such dynamics, state tracking and parameter estimation for a neural system fails. A beautiful overview of sub-grid effects in data assimilation is found in Kalnay's 2003 book. I would very much value the authors comments on these issues. I guess were I to respond to this question, that I might consider extending the model error term to an autoregressive model. But I will let the authors answer this question.

	\emph{The authors believe that allowing for time-varying parameters will account for changes in excitability levels. In future work, we will also estimate the disturbance properties (spatial covariance and temporal variance). To address this point in this paper, we have added an extra paragraph to `Neural Field Model' section of the Methods, where we discuss the model mismatch and some basic modeling assumptions, such as stationarity of the parameters. We have also added to the Estimation section, stating that we are assuming the parameters are stationary of the estimation period. In addition, we have added to the discussion (Extensions section) with possible ways to relax these assumptions. Model better estimation harder.}

### 2011-01-18 ltx/first_order_framework/response_to_reviews.tex

    \section{Reviewer 1}

    \subsection{Main Concerns}
    It would be useful to mention the above (now below) explicitly in the introduction and try to address them in later sections of the paper.

    \begin{enumerate}
        \item Do not use real data for the validation of their approach

	\emph{The authors appreciate that the only way to truly validate the proposed framework is with real data. However, before validating the method on real data we believe that it is essential to first validate the estimation method using accepted models and synthetic data, where the true and estimated parameters can be compared. In an effort to make the method employed in the paper more explicit, we have added ``To illustrate the estimation framework, data is generated using the neural field equations incorporating modeled sensors enabling a comparison between the estimated and true parameters'' to the introduction. We plan to validate the framework on real data in future work. We have added that validating the framework on real data is required in the Conclusion.}

@@

        \item Assume that a few kernel basis functions accurately represent the actual connectivity kernel, which might not be the case when using real data;

	\emph{The examples presented in the paper use the Mexican hat connectivity kernel. The framework is also valid for other connectivity kernels provided they are spatially homogeneous and that they can be approximated by a set of weighted isotropic basis functions. The mathematical formulation does not place any further restrictions on the form of the spatial mixing kernel. However, more complex shapes may require more basis functions, leading to more parameters. We have added statements to the methods section (after equation~17) to be more explicit in describing subtleties of the method.}

@@

	\emph{The authors believe that allowing for time-varying parameters will account for changes in excitability levels. In future work, we will also estimate the disturbance properties (spatial covariance and temporal variance). To address this point in this paper, we have added an extra paragraph to `Neural Field Model' section of the Methods, where we discuss the model mismatch and some basic modeling assumptions, such as stationarity of the parameters. We have also added to the Estimation section, stating that we are assuming the parameters are stationary of the estimation period. In addition, we have added to the discussion (Extensions section) with possible ways to relax these assumptions. Model better estimation harder.}

\item Although we all know the 1970s version of Wilson and Cowan's field equations, there was an intriguing stochasticization of these equations by Benayoun et al PloS Comp Biol 2010. I and others will value the authors' thoughts of how a more palatable stochasticization of these field equations might influence the future development of the data-assimilation framework presented in this paper.

\emph{In all honesty, the authors are not 100\% sure on what the reviewer is asking for with a more palatable stochasticization. We have attempted to answer assuming the reviewer is after a sensible method for choosing the disturbance parameters.}

\emph{The results of Benayoun et al. provide evidence that the variance in a stochastic WC model will depend on the number of neurons that are modeled (in a stochastic network), where a larger network behaves more like the deterministic WC equations. This work encourages the use of the WC model with an additive disturbance for data-assimilation, but provides evidence that the disturbance properties should also be estimated. The authors believe that estimating the noise properties is possible within an EM framework, and we are planning to demonstrate how to achieve this in a subsequent paper. We have added some comments to the Extension to the Framework section in this regard.}

### 2011-01-18 ltx/first_order_framework/LaTeX_Manuscript.tex
corrected equation C1 for Galerkein projection
edits to letter with suggestion from VK and Parham
edits to Euler appendix

% \newcommand{\dean}[1]{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}}
% \newcommand{\dean}[1]{\textcolor{red}{#1}}
% \newcommand{\parham}[1]{\textcolor{blue}{#1}}
% \newcommand{\mike}[1]{\textcolor{Plum}{#1}}

@@

The additive form of the unscented Rauch-Tung-Striebel smoother (URTSS) \citep{Sarkka2010} was used for the state estimation. The URTSS incorporates an unscented Kalman filter (UKF) \citep{Julier1997, Merwe2003} in a forward step to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward step to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. The sigma points, $\mathcal X_i$, are calculated using the unscented transform as follows:
\begin{align}\label{eq:sigmapoints1}
	\mathcal X_{0}&=\mathbf{\bar x} \\
	\mathcal X_{i}&= \mathbf{\bar x}+\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_i, \quad i=1, \dots, n_x \\
	\mathcal X_{i}&=\mathbf{\bar x}-\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_{i- n_x}, \quad i= n_x+1, \dots, 2n_x 
\end{align}
where $\mathbf{\bar x}$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$, $\mathbf{P}_x$ is the corresponding covariance matrix for filtering or smoothing, $\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_i$ is the $i^{th}$ column of the scaled matrix square root of $\mathbf P_x$, and $n_x$ is the dimension of the state space. The total number of sigma points is $2n_x+1$. The scaling parameter, $\lambda$, is defined as 
\begin{equation}\label{eq:sigmapoints3}
	\lambda=\alpha^2( n_x+\kappa) - n_x, 
\end{equation}
The positive constant $\alpha$, that determines the spread of the sigma points around $\mathbf{\bar x}$, was set to $10^{-3}$. It is typically set arbitrary small to minimise higher-order effects \citep{Haykin2001}. The other scaling parameter, $\kappa$, was set to $3-n_x$ in accordance to \citet{Julier2002a}.

@@

Table~\ref{tab:GaussianKernelResults} shows the statistics of the parameter estimates. Although the net reconstructed kernel estimates are in good accordance with the actual kernels, the individual kernel basis function weights are not in agreement with the actual parameters. This result demonstrates that the interpretation of the results should be based on the net shape of the estimated connectivity kernel and not on the estimated parameter values individually. These results indicate that the solution to the system may not be unique in non-ideal, practical situations. Nevertheless, the system is identifiable in terms of the net shape of the estimated kernel.

@@

If $0 < T_s < \tau$, where the synaptic time constant $\tau = 1/\zeta$, the IDE field model is stable. To show this, the model is written as the series
\begin{align}
	v_1(\mathbf{r}) &= \xi v_0(\mathbf{r}) + T_s g_0(\mathbf{r}) \\
	v_2(\mathbf{r})	&= \xi^2 v_0(\mathbf{r}) + T_s \left(\xi g_0(\mathbf{r}) + g_1(\mathbf{r}) \right) \\
	\vdots \, \, \, \, \, &= \, \, \, \, \, \vdots \nonumber \\
	v_{t+1}(\mathbf r) &= \xi^{t+1} v_0\left(\mathbf{r} \right) + T_s \sum_{i=0}^t \xi^i g_{t-i}(\mathbf{r}), \label{eq:fieldmodelseries}
\end{align}
where
\begin{equation}
	g_{t}(\mathbf{r}) = \int_\Omega { w\left(\mathbf{r}-\mathbf{r}'\right) f\left(v_t\left(\mathbf{r}'\right)\right) d\mathbf{r}'}.
\end{equation}
Since  $0 \leq f(v_t(\mathbf r)) \leq 1$ and the kernel is semi-compactly supported, $g_t(\mathbf r)$ is always bounded . To assess the stability of the system we assume that $g_t(\mathbf{r})$ is maximally saturated over the entire field for all time, which we denote as $g_{max}$. Substituting this into equation~\ref{eq:fieldmodelseries} yields
\begin{align}\label{vrecursiveinlimit}
	v_{t+1}(\mathbf r) &= \xi^{t+1}v_0(\mathbf r)+ g_{max} T_s \sum_{i=0}^t \xi^i \nonumber \\
	&=\xi^{t+1}v_0(\mathbf r) + \frac{ g_{max} T_s(1-\xi^{t+1})}{1-\xi}.
\end{align}
To study the asymptotic behaviour of the system we consider the limit as $t$ approaches infinity. Since $0 < \xi <1$  we use a property of the geometric series to simplify equation \eqref{vrecursiveinlimit} to
\begin{equation}\label{simplifiedvrecursiveinlimit}
	v_{t+1}(\mathbf r)=\xi^{t+1}v_0(\mathbf r)+\frac{g_{max} T_s}{1-\xi}.
\end{equation}
Substituting $\xi=1-\zeta T_s$ into equation \eqref{simplifiedvrecursiveinlimit} gives
\begin{align}
	v_{t+1}(\mathbf r) &= (1-\zeta T_s)^{t+1}v_0(\mathbf r)+\frac{g_{max}}{\zeta} \\
	&= \left(1-\frac{T_s}{\tau}\right)^{t+1}v_0(\mathbf r) + \tau g_{max}
\end{align}
From this equation it can be seen that if $0 < T_s < \tau$, the system is stable. The sampling period was set to $T_s = 0.001 s$ and the time constant was set to $\tau = 0.01 s^{-1}$. Therefore, the discretization preserved the stability properties of the system. Further analysis of the error introduced by Euler discretization method can be found in \citet{Butcher2008}.

@@

\section{Equivalence to Galerkin's Projection}\label{App:Galerkin}
To show how equation~\ref{eq:decompRHS} is equivalent to the Galerkin projection, we define the residual between the field and the decomposition as 
\begin{align}\label{eq:BasisFunctionResidual}
	\Xi(\mathbf{x}_t,v_{t+1}&(\mathbf{r})) = v_{t+1}(\mathbf{r}) \nonumber \\
	&- T_s\int_\Omega{f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t )\boldsymbol{\psi}^{\top}(\mathbf{r}-\mathbf{r}') \, d\mathbf{r}'}\boldsymbol{\theta} \nonumber \\
	&+ \xi\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_t + \mathbf{e}_t(\mathbf{r}).
\end{align} 
The Galerkin projection finds $\mathbf{x}_t$, such that the inner product of the residual with all of the basis functions satisfies
\begin{equation}\label{eq:GalerkinProjection}
	G_i = \int_{\Omega}\Xi(\mathbf{x}_t,v_{t+1}(\mathbf{r})) \phi_i(\mathbf{r}) d\mathbf{r} = 0,
\end{equation}
where $i=0,1,\hdots,n_x-1,$ and $n_x$ is the number of states and basis functions. If equation~\ref{eq:GalerkinProjection} is satisfied the basis functions are orthogonal to the residual. By substituting equation~\ref{eq:BasisFunctionResidual} into equation~\ref{eq:GalerkinProjection} and rearranging we get equation~\ref{eq:decompRHS} of the main text. Note, if Galerkin's method was used to estimate the states, $n_x$ coupled ODEs would need to be solved independently (each $G_i$) to estimate the state vector, which is a standard method in fluid dynamics and economics modeling. The method in this paper takes a different approach by forming the state-space model, allowing for more efficient estimation.

>>>>

\newcommand{\dean}[1]{\textcolor{red}{#1}}
\newcommand{\parham}[1]{\textcolor{blue}{#1}}
% \newcommand{\mike}[1]{\textcolor{Plum}{#1}}

@@

The additive form of the unscented Rauch-Tung-Striebel smoother (URTSS) \citep{Sarkka2010} was used for the state estimation. The URTSS incorporates an unscented Kalman filter (UKF) \citep{Julier1997, Merwe2003} in a forward step to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward step to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. The sigma points, $\mathcal X_i$, are calculated using the unscented transform as follows:
\begin{align}\label{eq:sigmapoints1}
	\mathcal X_{0}&=\mathbf{\bar x} \\
	\mathcal X_{i}&= \mathbf{\bar x}+\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_i, \quad i=1, \dots, n_x \\
	\mathcal X_{i}&=\mathbf{\bar x}-\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_{i- n_x}, \quad i= n_x+1, \dots, 2n_x 
\end{align}
where $\mathbf{\bar x}$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$, $\mathbf{P}_x$ is the corresponding covariance matrix for filtering or smoothing, $\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_i$ is the $i^{th}$ column of the scaled matrix square root of $\mathbf P_x$, and $n_x$ is the dimension of the state space. The total number of sigma points is $2n_x+1$. The scaling parameter, $\lambda$, is defined as 
\begin{equation}\label{eq:sigmapoints3}
	\lambda=\alpha^2( n_x+\kappa) - n_x, 
\end{equation}
The positive constant $\alpha$, that determines the spread of the sigma points around $\mathbf{\bar x}$, was set to $10^{-3}$. It is typically set arbitrary small to minimise higher-order effects \citep{Haykin2001}. The other scaling parameter, $\kappa$, was set to $3-n_x$ in accordance to \citet{Julier2002a}. \dean{Parham to add something about the sensitivity of the parameters.}

@@

Table~\ref{tab:GaussianKernelResults} shows the statistics of the parameter estimates. Although the net reconstructed kernel estimates are in good accordance with the actual kernels, the individual kernel basis function weights are not in agreement with the actual parameters. This result demonstrates that the interpretation of the results should be based on the net shape of the estimated connectivity kernel and not on the estimated parameter values individually. These results indicate that the solution to the system may not be unique in non-ideal, practical situations. Nevertheless, the net shape of the connectivity kernel can be determined reasonably accurately.

@@

% If $0 < T_s < \tau$, where the synaptic time constant $\tau = 1/\zeta$, the IDE field model is stable. To show this, the model is written as the series
% \begin{align}
% 	v_1(\mathbf{r}) &= \xi v_0(\mathbf{r}) + T_s g_0(\mathbf{r}) \\
% 	v_2(\mathbf{r})	&= \xi^2 v_0(\mathbf{r}) + T_s \left(\xi g_0(\mathbf{r}) + g_1(\mathbf{r}) \right) \\
% 	\vdots \, \, \, \, \, &= \, \, \, \, \, \vdots \nonumber \\
% 	v_{t+1}(\mathbf r) &= \xi^{t+1} v_0\left(\mathbf{r} \right) + T_s \sum_{i=0}^t \xi^i g_{t-i}(\mathbf{r}), \label{eq:fieldmodelseries}
% \end{align}
% where
% \begin{equation}
% 	g_{t}(\mathbf{r}) = \int_\Omega { w\left(\mathbf{r}-\mathbf{r}'\right) f\left(v_t\left(\mathbf{r}'\right)\right) d\mathbf{r}'}.
% \end{equation}
% Since  $0 \leq f(v_t(\mathbf r)) \leq 1$ and the kernel is semi-compactly supported, $g_t(\mathbf r)$ is always bounded . To assess the stability of the system we assume that $g_t(\mathbf{r})$ is maximally saturated over the entire field for all time, which we denote as $g_{max}$. Substituting this into equation~\ref{eq:fieldmodelseries} yields
% \begin{align}\label{vrecursiveinlimit}
% 	v_{t+1}(\mathbf r) &= \xi^{t+1}v_0(\mathbf r)+ g_{max} T_s \sum_{i=0}^t \xi^i \nonumber \\
% 	&=\xi^{t+1}v_0(\mathbf r) + \frac{ g_{max} T_s(1-\xi^{t+1})}{1-\xi}.
% \end{align}
% To study the asymptotic behaviour of the system we consider the limit as $t$ approaches infinity. Since $0 < \xi <1$  we use a property of the geometric series to simplify equation \eqref{vrecursiveinlimit} to
% \begin{equation}\label{simplifiedvrecursiveinlimit}
% 	v_{t+1}(\mathbf r)=\xi^{t+1}v_0(\mathbf r)+\frac{g_{max} T_s}{1-\xi}.
% \end{equation}
% Substituting $\xi=1-\zeta T_s$ into equation \eqref{simplifiedvrecursiveinlimit} gives
% \begin{align}
% 	v_{t+1}(\mathbf r) &= (1-\zeta T_s)^{t+1}v_0(\mathbf r)+\frac{g_{max}}{\zeta} \\
% 	&= \left(1-\frac{T_s}{\tau}\right)^{t+1}v_0(\mathbf r) + \tau g_{max}
% \end{align}
% From this equation it can be seen that if $0 < T_s < \tau$, the system is stable. The sampling period was set to $T_s = 0.001 s$ and the time constant was set to $\tau = 0.01 s^{-1}$. Therefore, the discretization preserved the stability properties of the system. 
Note, the discretization using the first Euler method has the potential to affect the system's behaviour. If the sampling period is too large, the difference between the continuous and discrete-time systems can be substantial. The sampling period of 1~ms was used in this paper, which was sufficiently small to yield good estimation results. An analysis of the error introduced by Euler discretization method can be found in \citet{Butcher2008}. An alternative to the Euler's method is the local linearization method, which may provide improved results~\citep{Ozaki1994}.

@@

\section{Equivalence to Galerkin's Projection}\label{App:Galerkin}
To show how equation~\ref{eq:decompRHS} is equivalent to the Galerkin projection, we define the residual between the field and the decomposition as 
\begin{align}\label{eq:BasisFunctionResidual}
	\Xi(\mathbf{x}_t,v_{t+1}&(\mathbf{r})) = v_{t+1}(\mathbf{r}) \nonumber \\
	&- T_s\int_\Omega{f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t )\boldsymbol{\psi}^{\top}(\mathbf{r}-\mathbf{r}') \, d\mathbf{r}'}\boldsymbol{\theta} \nonumber \\
	&- \xi\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_t - \mathbf{e}_t(\mathbf{r}).
\end{align} 
The Galerkin projection finds $\mathbf{x}_t$, such that the inner product of the residual with all of the basis functions satisfies
\begin{equation}\label{eq:GalerkinProjection}
	G_i = \int_{\Omega}\Xi(\mathbf{x}_t,v_{t+1}(\mathbf{r})) \phi_i(\mathbf{r}) d\mathbf{r} = 0,
\end{equation}
where $i=0,1,\hdots,n_x-1,$ and $n_x$ is the number of states and basis functions. If equation~\ref{eq:GalerkinProjection} is satisfied the basis functions are orthogonal to the residual. By substituting equation~\ref{eq:BasisFunctionResidual} into equation~\ref{eq:GalerkinProjection} and rearranging we get equation~\ref{eq:decompRHS} of the main text. Note, if Galerkin's method was used to estimate the states, $n_x$ coupled ODEs would need to be solved independently (each $G_i$) to estimate the state vector, which is a standard method in fluid dynamics and economics modeling. The method in this paper takes a different approach by forming the state-space model, allowing for more efficient estimation.
